\documentclass[12pt]{article}

% House style preamble
\input{.house-style/preamble.tex}

% Update PDF metadata
\hypersetup{
  pdftitle={What do we mean by language}
}

\title{What do we mean by \mention{language}? \\
  \large A pluralist map for the language sciences}
\author{Brett Reynolds \orcidlink{0000-0003-0073-7195}%
\thanks{Contact: \href{mailto:brett.reynolds@humber.ca}{brett.reynolds@humber.ca}}\\
Humber Polytechnic \& University of Toronto}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
The word \mention{language} names several objects in the language sciences: cognitive capacity, individual command, communicative practice, population convention, named sociohistorical object, and formal system. Existing pluralist accounts, especially Santana's distinction among psychological, social, and abstract ontologies, show why monism fails, but they don't explain when a proposed sense licenses inference or when evidence about one sense can travel to another. This paper adds a homeostatic-property-cluster discipline. A language sense earns realist standing where its properties are stabilized by identifiable mechanisms and project for a field purpose; where the mechanisms diverge, the shared label doesn't license transfer. The resulting map treats capacity, idiolect, and communicative practice as neighbourhoods rather than unified categories, and treats population convention, named sociohistorical object, and abstract formal system as categories with different stabilizers and projection targets. The payoff is diagnostic. The Casten / Pinker exchange over genomic evidence for spoken-language ability shows how developmental verbal-performance evidence can be read too quickly as evidence about a uniquely human language faculty. Similar slippages arise when census recognition is read as convention, formal models as cognitive architecture, or analyst diagnostics as the object itself. The umbrella term remains useful for coordination, but the load-bearing scientific units are the categories and the neighbourhoods, not the umbrella.
\end{abstract}

\section{A slippage and a question}\label{sec:intro}

In April 2026, \textcite{casten2026} reported that a small fraction of the human genome carries an outsized share of variation in spoken language ability. Their target measure, the F1 factor, is anchored by sentence repetition and aggregates morphosyntactic, phonological, working-memory, lexical, and motor contributions to verbal performance. In a community-based cohort of 350 children assessed longitudinally from kindergarten through fourth grade, a polygenic score restricted to human-ancestor quickly-evolved regions (HAQERs, less than 0.1\% of the genome) showed an effect of $\beta = 0.18$ on F1, no effect on nonverbal IQ, and replication across more than 100,000 participants in three further cohorts with phenotypically coarser measures. Cross-species analysis of 170 nonprimate mammalian genomes found that HAQER-like sequence similarity correlates with vocal-learning ability across 49 vocal-learning species and 121 non-vocal learners.

Three days later, on 25 April 2026, Steven Pinker cited the paper on X \citep{pinker2026tweet} as vindication of \textcite{pinkerbloom1990}. The Pinker--Bloom thesis, as Pinker summarizes it on the day, is that the human language faculty is \enquote{a genetically complex trait, distinct from general cognition, which evolved gradually by natural selection of many regulatory genes over a long period.} The Casten paper's polygenic-score, ancient-DNA, and selection-signal results, Pinker writes, \enquote{suggest we were right.} The reading was widely circulated and approvingly relayed. It is also a slippage.

What \citeauthor{casten2026} measured is a developmental verbal-performance phenotype anchored by sentence repetition: a score that loads on attention, working memory, and lexical access alongside grammatical processing. The HAQER-like regulatory regions they identify are shared across nonprimate mammalian vocal learners (cetaceans, pinnipeds, and bats), not human-specific. Pinker--Bloom's faculty is supposed to be \emph{distinct from general cognition} and human-specific; the regulatory architecture Casten et al.\ identify is shared with nonprimate vocal learners and underwrites a developmental phenotype loading on multiple cognitive substrates. Neither finding bears directly on the architecture of a uniquely human grammatical capacity. The shared label \mention{language} obscures the move. A developmental verbal-performance phenotype, anchored by regulatory architecture humans share with vocal-learning mammals, gets read as evidence about the human-specific complex faculty because both are referred to as evidence about \mention{language}.

This isn't a one-off. The word \mention{language} does work for several recurrent senses across linguistics, cognitive science, anthropology, philosophy, semiotics, law, and education. It picks out an individual mental system in one paragraph and a community-level convention in the next; a developmental phenotype here and an institutionally-stabilized national object there; a set of formal strings in semantic theory and a situated communicative practice in linguistic anthropology. Each of these senses is real, well-studied, and distinct in its projection target. None licenses inferences about the others without further argument. Treating evidence about one as evidence about another is the recurrent pattern that the Casten / Pinker case shows in miniature.

The question of this paper, then, is how we should use the word \mention{language} across fields, given that the word does work for senses that don't all license the same inferences? Two answers already on offer fall short. The first is monism: pick one sense and treat the rest as misapplications. \textcite[26]{Chomsky1986} adopts this stance for psychological language and dismisses social senses as \enquote{artificial, somewhat arbitrary, and perhaps not very interesting constructs}. \textcite{santana2016language} shows the dismissal fails: positive arguments for each sense succeed, and the negative arguments for ruling out the others generally fail. The second is unconstrained pluralism: accept the senses as a list and leave it at that. Santana himself rules this out -- the senses interact, and confirmation, idealization, and theory choice all turn on which sense a researcher is committed to.

This paper takes a third route. The senses cluster: they are recurrent, recognisable, and rough enough at the edges that working scientists already treat them as different. Each earns realist standing \emph{for the field purposes its mechanisms underwrite} (with the further refinement, developed in §\ref{sec:pluralism}, that a putative cluster whose sub-divisions don't share a stabilizer set is best diagnosed as a neighbourhood whose sub-categories project independently). When a claim asks one to project beyond that range, the realist commitment doesn't disappear, but the inference needs an additional bridge. Boyd's homeostatic-property-cluster framework (categories as profiles of co-occurring properties, stabilized by mechanisms, projectible relative to purposes; \citealt{boyd1991,boyd1999}) supplies the discipline. The contribution is the synthesis: a more granular taxonomy than Santana's three ontologies, plus the diagnostic that lets us say when claims about one element smuggle in claims about another. The Casten / Pinker case is a worked instance of exactly that diagnosis. Section~\ref{sec:slippage} works through it in detail.

The map proposed has six elements: three neighbourhoods and three categories. The three neighbourhoods are capacity (related sub-categories at the individual-internal scale, sharing a research community but not a unifying stabilizer set), idiolect (gathering several ontological positions about the individual-level linguistic entity that don't share a stabilizer set either), and communicative practice (sub-categories drawn from speech-act theory, interactional sociolinguistics, communicative-competence theory, and metapragmatic typification, sharing a research community but projecting to non-subsuming targets). The three categories are population-level convention, named sociohistorical object, and abstract formal system. The criterion for distinguishing the cases, set out in §\ref{sec:pluralism}, runs through shared stabilizers, projection-target overlap, and perturbation coupling. The taxonomy is open: new categories or neighbourhoods are admissible if they pass the projectibility and homeostasis tests. Section~\ref{sec:umbrella} returns to why the umbrella label \mention{language} remains useful as a coordinative term across fields. The elements under it are what working science actually projects from.

\section{Pluralism, and what HPC adds}\label{sec:pluralism}

The closest extant precedent for what follows is \textcite{santana2016language}. Santana inventories the candidate ontologies that linguists work with~-- psychological (Chomskyan i-language; psycholinguistic processing), social (Lewis-style convention; Labov's speech communities; Devitt's externalism), and abstract (Katz's Platonism; Montague semantics; formal-language theory)~-- and shows that the positive arguments for each succeed while the negative arguments for ruling out the others generally fail. The result is a pluralist ontology of language: more than one of these classes is real, and which one a researcher's claims target depends on what kind of inferences they're trying to license.

Santana frames this as three criteria any candidate ontology has to meet. The candidate has to exist in a form accessible to scientific study, it has to be a primary object of study for some subfield in practice, and it has to be reasonably called \mention{language}. He then attaches two strictures to keep the pluralism honest. First, no isolationism: pluralism doesn't license treating the three classes as three Kuhnian paradigms with nothing to say to each other. Research on each ontology bears indirectly on the others, and the relations among them are themselves worth investigating. Second, no agnosticism: in any particular research context, the scientist has to be explicit about which ontology they're committed to, because confirmation, idealization, and theory choice all turn on that commitment.

Santana's framework gets the lay of the land right. What it doesn't supply is a discipline for testing whether a given category earns its realist standing in the work it's asked to do. Three things aren't yet in hand. The first is a test for projectibility: a category has to license inferences from observed members to unobserved ones, or it isn't earning its keep. The second is a test for homeostasis: the category has to be sustained by identifiable mechanisms whose perturbation predicts patterned decay, or the apparent clustering is a labelling convenience. The third is a way of diagnosing slippage: when a researcher takes evidence about one category as evidence about another because both are called \mention{language}, the pattern has to be visible as a methodological failure when no bridging argument is supplied, rather than as an unmarked rhetorical move.

\textcite{boyd1991, boyd1999} supplies the missing piece. The homeostatic-property-cluster framework was developed for biological species and other natural kinds whose members share a profile of co-occurring properties without a defining essence. A \term{cluster} is a profile of properties that tend to go together. A cluster earns \term{category} status, in the realist sense needed here, when that profile projects: learning about observed cases supports counterfactual expectations about unobserved ones. Three checks discipline the claim. The cluster check asks whether property covariance holds. The homeostasis check asks whether identifiable stabilizers explain that covariance and predict patterned change when perturbed. The projectibility check asks whether the cluster supports the inferences a field wants to draw from it. Passing the checks for one field's purposes doesn't entail passing them for another \citep[see also][on cluster categories for cognitive and human kinds]{khalidi2013}. \term{Field-relative projectibility} is the Boyd move that lets two fields carve overlapping extensions differently, with each carving disciplined by what mechanisms actually do; the \mention{Cantonese}-as-named-object case in §\ref{sec:slippage} works the asymmetry concretely.

The framework applies recursively. At any level, a cluster either projects as one category or it doesn't, with projection then living at finer grain among constituent sub-categories. Three diagnostics distinguish single-category status from a coordinative label organizing several sub-categories. First, do the sub-divisions share a stabilizer set, or do they have different stabilizing mechanisms at different grains? Second, do they license overlapping inferences, or do their projection targets fail to subsume one another? Third, does perturbing one sub-division predict patterned decay across the others, or are they decoupled? When the candidate's sub-divisions answer no to all three, the named label is coordinative rather than a single category. Coordinative labels come in two degrees of coupling. A \term{neighbourhood} has substantively coupled sub-categories: shared scale of inquiry, shared explanandum, and mechanistic, developmental, or structural relations among them. Capacity's sub-categories share individual-internal cognitive and biological architecture; idiolect's share the individual-scale explanation of what speakers know; communicative practice's share situated joint-action as their object. An \term{umbrella} has loosely coupled elements: shared field of inquiry and research community, but different scales, different mechanism types, no shared explanandum. The umbrella label \mention{language}, treated in §\ref{sec:umbrella}, organizes capacity-research alongside formal-language theory and population-convention work, across scales and mechanism types that don't share a target. \citeauthor{khalidi2013}'s permissive line would call neighbourhoods themselves cluster categories at coarser grain; this paper adopts the stricter Boyd-style line, where coordinative labels organize categories without themselves projecting as one.

The verdicts are current-state and revisable in both directions. A neighbourhood (or umbrella) can be promoted to single-category status if a unifying stabilizer set is later demonstrated, as scientific unification has historically merged what looked like separate kinds (Maxwell's electromagnetism over electrostatic and magnetic phenomena; the modern synthesis over Mendelian inheritance and Darwinian selection). A category can be split or demoted to a coordinative label if closer inspection shows its sub-divisions don't share the unifying mechanism the category status assumed. The sections below apply the test where the answer isn't obvious; capacity (§\ref{sec:capacity}), idiolect (§\ref{sec:idiolect}), and communicative practice (§\ref{sec:communicative}) come out as neighbourhoods on the present evidence, while population-level convention (§\ref{sec:population}), named sociohistorical objects (§\ref{sec:named}), and abstract formal systems (§\ref{sec:formal}) satisfy single-category status.

The contribution of this paper is the synthesis. The taxonomy below is finer-grained than Santana's three classes: each of the social, psychological, and abstract classes contains more than one category, with some passing single-category status and some resolving into a neighbourhood of related sub-categories with distinct projection targets and stabilizers. The HPC discipline supplies what Santana's pluralism leaves undone. And \term{slippage} across categories (the recurring move where evidence licensed by the stabilizers of one element is read as evidence about another that the term \mention{language} also covers) becomes diagnosable in HPC terms, as a mismatch between the projection target a researcher's evidence actually licenses and the target their conclusion is reaching for.

A short example shows the subtraction at the level of inference. Santana's three criteria license treating \mention{English subject-verb agreement} as a real object of inquiry: it exists in usable form, has primacy across formal syntax, sociolinguistics, and corpus linguistics, and is reasonably called part of English. The two strictures hold too: claims about agreement bear on neighbouring features (no isolationism); researchers studying agreement commit to it as the object (no agnosticism). Santana's framework licenses claims of the form \enquote{English speakers display existential-\mention{there} agreement at rate R}. HPC's homeostasis check rules out a class of those claims. Section~\ref{sec:population}'s audit shows the move: the pooled rate across speakers conceals two register-conditioned subsystems whose stabilizers differ, and the property covariance under the pooled label fails to reconstitute under perturbation. Santana's framework permits the inference; HPC's discipline tells you which grain the inference actually projects from. The subtraction is what the discipline does over Santana on the same object, and §§\ref{sec:population}--\ref{sec:slippage} catalogue cases like this across the map.

The six elements all sit within a wider semiotic field that includes animal communication, programming languages, formal symbol systems, and mathematical notations. Treating these as boundary cases for the human-language elements keeps the inquiry calibrated; treating the wider genus as itself an HPC category over-extends the framework, since the elements that compose it are maintained by mechanisms too disparate to project as one. Animal communication and programming languages will appear here as boundary phenomena under capacity and communicative practice, not as elements of their own.

Each element of the map gets operationalized differently in research practice: corpora and treebanks for the social elements, competence-and-performance idealizations for capacity, formal-language-theoretic objects for the abstract-formal category, ethnographic and conversation-analytic transcription for communicative practice. Claims of the form \enquote{the corpus IS the language}, \enquote{competence IS the language}, \enquote{the treebank IS the grammar}, or \enquote{the LLM output IS the language} substitute the analyst's diagnostic for the element it's supposed to track. Section~\ref{sec:slippage} catalogues these alongside the across-the-map cases.

\section{Capacity}\label{sec:capacity}

\term{Capacity} is what the literature gathers about what humans can do with language: acquire, process, perceive, produce, and store linguistic material. \citeauthor{pinkerbloom1990} argue it's a complex adaptation \citep{pinkerbloom1990}; \textcite{hauser2002} partition it into broad and narrow faculties; developmental-genetic and clinical-genetic studies aim at it when they look for individual differences. Capacity isn't one category, though. The five sub-categories that recur in the literature have stabilizing mechanisms operating at different grains, license different inferences, and don't share a single homeostasis at the level the neighbourhood label implies. Capacity is best treated as a neighbourhood of related sub-categories at the individual-internal scale, sharing family resemblance but not a unifying stabilizer set.

The first sub-category is the abstract computational architecture: what is uniquely human and uniquely linguistic in the system that maps form onto meaning. \textcite{hauser2002} call this the faculty of language in the narrow sense (FLN) and propose that it consists of recursive computation plus the interfaces with sensory-motor and conceptual-intentional systems. \textcite{pinkerjackendoff2005} contest this characterization, arguing that phonology has discrete infinity but not technical recursion, that words encode language-specific syntactic information not captured by recursion alone, and that case, agreement, and most of the syntactic apparatus beyond hierarchical phrase grouping are language-specific. Stabilizers proposed for the architecture sub-category run from genetic endowment (Universal Grammar) through cognitive-architectural primitives like Merge to the convergent product of multiple selected components. Observed in this sub-category, P (a structural feature like binding or displacement) licenses expected Q (a typological universal or an acquisition-order prediction) in unobserved cases; defeated if D (the feature is acquired by general-cognitive mechanisms, varies cross-linguistically in ways UG-style theories can't accommodate, or appears outside language entirely).

The second sub-category is the broad faculty: the sensory-motor system, the conceptual-intentional system, and their interfaces with the architectural sub-category above. \citeauthor{hauser2002}'s claim is that most of the broad faculty is shared with other species, including capacities for vocal imitation, categorical perception of speech-like sounds, theory of mind, and conceptual representation of objects, agents, and events. The methodological backbone here is comparative biology. Observed P (a sensory-motor or conceptual capacity in species $X$) licenses expected Q (the same capacity in species $Y$, tracking shared inheritance) in unobserved cases through the comparative-method warrant: homology supported by phylogenetic continuity predicts trait recurrence under inherited stabilizers, while convergent evolution under similar selection pressures predicts recurrence under analogous-but-not-homologous stabilizers. Defeated if D: the capacity proves uniquely human (re-classifying it as FLN material), the comparative analogue fails under denser species sampling, or phylogenetic confounds drive the apparent cross-species correlation.

The third sub-category is developmental-acquisition. Children acquire languages at species-specific rates and modes, regardless of modality (spoken or signed), with sensitive-period effects on phonology, vocabulary, and morphosyntax. The Casten et al.\ F1 measure introduced in §\ref{sec:intro} sits squarely here: a longitudinal aggregate of sentence repetition, working memory, attention, and lexical retrieval scores from children sampled at kindergarten through fourth grade. Stabilizers are developmental input, neural and cognitive maturation, and modality-general abstraction. Observed P (an acquisition pattern in a cohort) licenses expected Q (a clinical or developmental outcome in unobserved children with similar input environments; or a replicable association with a specific genetic variant) in unobserved cases; defeated if D (replication fails, or the measure turns out to track general cognitive variance rather than language-specific developmental trajectory).

The fourth sub-category is comparative-biology of vocal learning. Convergent regulatory architecture for vocal learning shows up in genomic comparisons that control for phylogenetic relatedness. Within placental mammals, \textcite{wirthlin2024} report convergent regulatory specializations across the four mammalian vocal-learning lineages (bats, cetaceans, pinnipeds, and humans), drawing on 215 mammalian genomes. The HAQER-like sequences from \textcite{casten2026}, correlated with vocal-learning ability across 49 vocal-learning and 121 non-vocal-learner species at the cross-taxon scale, sit alongside. Stabilizers are convergent regulatory tuning, transcription-factor binding affinities, and selection on auditory-cortex temporal processing. Observed P (a regulatory architecture in species $X$) licenses expected Q (a similar architecture in species $Y$ conditional on convergent selection pressure) in unobserved cases; defeated if D (the convergence claim fails under denser sampling, or the correlation tracks a phylogenetic confound rather than vocal-learning function).

The fifth sub-category is impairment phenotype. Specific Language Impairment, FOXP2-related disorders, dyslexia, and developmental stuttering pick out variation that is heritable and partly localizable to specific genetic loci, but the impairments span multiple grammatical and processing components rather than knocking out a single circuit. The argument from impairment phenotype is one of \citeauthor{pinkerjackendoff2005}'s pieces of evidence against the recursion-only hypothesis: heritable language impairments aren't heritable recursion impairments. Stabilizers here are specific genetic loci with measurable effects, polygenic background, and developmental cascades from molecular variation to behavioural phenotype. Observed P (a specific genetic variant) licenses expected Q (a related deficit in unobserved cases sharing the variant) in unobserved cases; defeated if D (the impairment phenotype is mechanistically heterogeneous, or the variant's effect doesn't replicate under stricter assessment).

The sub-categories overlap, and a single study often touches more than one. They license different inferences, though, so claims travelling between them need explicit warrant. The Casten / Pinker case from §\ref{sec:intro} is exactly such a traversal: a developmental-acquisition phenotype anchored by a comparative-biology regulatory architecture, read as evidence about the abstract computational architecture. The HCF / PJ exchange is a similar dance about which sub-category the FLN claim actually targets.

Treating capacity as a neighbourhood rather than a single category does two things a Santana-style class-of-the-psychological treatment doesn't. First, the sub-category individuation makes traversals across sub-categories visible as moves that need warrant: \citeauthor{casten2026} measure sub-categories 3 and 4; \citeauthor{pinkerbloom1990}'s faculty argument lives at sub-category 1; the evidence in hand doesn't entitle the inference from one to the other. Second, projection targets get specified at the sub-category grain where the relevant stabilizers actually do their work, not at the level of the neighbourhood label that bundles distinct mechanism sets. Section~\ref{sec:slippage} works through the Casten / Pinker case in detail.

\section{Idiolect and \mentionhead{I-language}}\label{sec:idiolect}

The grouping around the word \mention{idiolect} (and Chomsky's \mention{I-language}) covers \mention{language} claims at the individual scale: what a speaker knows, commands, or has internalized. Applied to this grouping, the test from §\ref{sec:pluralism} returns a neighbourhood verdict rather than a single category. Sub-category~1 (Chomskyan I-language) and sub-category~2 (Devittian idiolect) share a label and a research community but disagree at the ontological level about what the individual-level linguistic entity actually is, and the disagreement runs through to stabilizers: I-language stabilizers are genetic-cognitive (UG plus maturation), Devittian stabilizers are convention-based (Lewis-style mutual expectation plus individual entrenchment), and sub-category~3's empirical-individual-variation stabilizers are exposure-history-and-network-density. Projection targets diverge correspondingly. Perturbing individual exposure (sub-category~3) doesn't predict patterned decay across the architectural sub-category~1 or the convention sub-category~2. The grouping is best treated as a neighbourhood of related sub-categories at the individual scale.

The first sub-category is \mention{I-language} in \citeauthor{Chomsky1986}'s sense: the mental system in a competent speaker, the steady state reached at the end of acquisition, and the species-uniformly-shared universal grammar that constrains which I-languages are humanly possible. Stabilizers proposed for I-language run from genetic endowment to maturational mechanisms to consolidation in adulthood, with linguistic input from the community as the trigger. The minimalist programme's Merge sits here as a candidate computational primitive. Observed P (an acceptability-judgement asymmetry across sentence types, an acquisition-order pattern in one cohort, an aphasia profile in one patient population) licenses expected Q (the same asymmetry replicated in typologically-related languages following UG-constraint-rank ordering, the same acquisition order in new cohorts, the same selective-sparing pattern in new aphasic patients) in unobserved cases; defeated if D (the same P pattern is derivable from non-mental mechanisms, the predicted Q fails to replicate, or alternative frameworks constrain it as tightly).

The second sub-category is the idiolect as \textcite{devitt2008} characterizes it: the individual's command over a non-psychological realm of physical linguistic entities~-- sounds, inscriptions, gestures~-- whose linguistic properties (syntactic, semantic, phonological) are conventionally established. Devitt's argument turns on three distinctions: between a theory of competence and a theory of its outputs, between structure rules governing the outputs and processing rules governing the exercise, and between a competence respecting the structure rules and one including those rules among its processing rules. The grammar, on this view, is a theory of the linguistic reality, and what it tells us about the speaker's mind is just that the speaker has \enquote{something-we-know-not-what} that respects the rules \citep[219]{devitt2008}. Stabilizers are convention (Devitt invokes \citealt{lewis1969} explicitly), entrenchment of conventional uses, and mutual understanding within speech communities. Observed P (a structural property of an expression class) licenses expected Q (the same property in unobserved expressions of similar type) in unobserved cases through Lewis-style convention extension: speakers' mutual expectation of the conventional property carries the inference to novel cases of similar structure. Perturbing the convention (declining mutual exposure, register shift, novel coinage outside the convention's reach) predicts patterned decay in P--Q reliability. Defeated if D: the conventional properties don't actually hold for the predicted cases (the convention has fragmented), or the conventions can't be established for the relevant cases (the population is too thinly connected for common knowledge).

The third sub-category is empirical individual variation: differential competence across individuals, in Hymes's heterogeneous-speech-community sense \citep{hymes1972}. Bloomfield's contrast between White Thunder, whose Menominee was thin and acculturation-shaped, and Red Cloud Woman, who commanded Menominee, Ojibwa, and Potawatomi fluently, sits here. D\k{a}browska's documentation of substantial individual variation in grammatical knowledge (in rule extraction, morphological productivity, and comprehension of complex constructions, with education and written-language exposure as recurring predictors) sits here too \citep{dabrowska2012}. Stabilizers are acquisition history, exposure-driven entrenchment, and individual network density. Observed P (a single speaker's variation pattern) licenses expected Q (similar patterns in similar individuals; predictable variation under conditioning on input history) in unobserved cases; defeated if D (the variation is uniformly random, or is fully explained by population-level convention without residual individual signal).

The Devitt--Chomsky exchange that animates this neighbourhood is itself sub-category distinction work being negotiated in real time. Chomsky takes the individual-level entity to be a mental system; Devitt takes it to be the individual's command over public-language expressions; \textcite{collins2007, collins2008} press Devitt on whether there's a non-psychological linguistic reality to study at all; Rey's antirealist argument denies that the relevant linguistic entities exist in either form \citep{rey2006a, rey2006b}. The HPC discipline doesn't adjudicate the ontological dispute, but it does locate where each disputant's claims project. Sub-category 1 projects to UG-shaped predictions about cross-linguistic stability and maturational trajectory; sub-category 2 projects to convention-extension predictions about novel expressions; sub-category 3 projects to variation patterns under input perturbation. Treating data licensed by one sub-category as direct evidence for another is the within-neighbourhood slippage diagnosed in §\ref{sec:slippage}.

\section{Communicative practice}\label{sec:communicative}

The grouping of \mention{language} claims about \emph{what speakers are doing when they communicate} is not a residual catch-all. It carries its own analytical engine, drawn from speech-act theory, conversation analysis, interactional sociolinguistics, and linguistic anthropology, and the engine handles phenomena (conventionalized illocutionary force, interactional uptake, situated competence, register formation, metapragmatic typification) that no other element of the map captures at the right grain. Applied here, the §\ref{sec:pluralism} test returns a neighbourhood verdict. Stabilizers vary across sub-categories: constitutive rules and felicity conditions for speech acts; alignment and turn-taking machinery for interactional organization; conditioned-by-context appropriateness for communicative competence; reflexive enregisterment for metapragmatic typification. Projection targets fail to subsume one another. Predicting interactional uptake from the conversation-analytic apparatus doesn't entail predicting register inference from the metapragmatic apparatus, and vice versa. Perturbing one sub-category doesn't predict patterned decay across the others: enregisterment can fragment without felicity conditions for promising shifting, and felicity conditions for declarations can shift (the legal performative reform case) without conversation-analytic turn-taking machinery breaking. The grouping shares a research community and an analytical engine, but the homeostasis is plural rather than unified. Treating it as a neighbourhood at the situated-and-conventional scale, parallel to capacity at the individual-internal scale, makes within-grouping traversals legible as moves needing warrant.

The first sub-category is speech acts. \textcite{austin1962} and \textcite{searle1969} showed that uttering can be doing (promising, requesting, asserting, declaring, apologizing) and that the doing is conventional rather than expressive. The illocutionary force of an utterance is constituted by rules at the population level, with felicity conditions specifying what counts as a successful instance. Stabilizers are conventionalized constitutive rules, acquisition of speech-act conventions in the same developmental matrix as grammar, and institutional infrastructure for high-stakes speech-act types (legal performatives, religious oaths, ceremonial declarations). Observed P (a form deployed in context X by a speaker oriented to community C's speech-act conventions) licenses expected Q (predictable uptake by interlocutors oriented to C; predictable repair when the form fails to coordinate; the same form being performed-as the speech-act type in unobserved exchanges within C; cross-community variation in which forms perform which acts) in unobserved cases; defeated if D (the form fails to count as the speech-act type for participants in the actual interaction~-- mis-coordination, cultural difference in conventions, ironic or mock deployment).

The second sub-category is interactional and situational organization. \textcite{goffman1981} on footing and participation frameworks, conversation-analytic work on turn-taking and repair \citep{SacksSchegloff1979, Schegloff1977}, and interactive-alignment research on dyadic coordination \citep{PickeringGarrod2004} together specify how interaction is sequentially managed: who speaks when, how repair is initiated and completed, how participation roles (speaker, addressee, overhearer, ratified third party) are shifted by footing, and how utterances are understood as moves in a sequentially-organized exchange. \emph{Uptake} is the load-bearing notion: what an utterance does in the interaction depends on how subsequent turns treat it. Stabilizers are interactive alignment at the dyadic level, cultural transmission of interactional conventions, and institutional gatekeeping in high-stakes contexts (courtroom, classroom, medical encounter). Observed P (an interactional move or sequence in context X) licenses expected Q (predictable response from interlocutor; predictable repair if the move violates a norm; predictable participation-framework consequences) in unobserved cases; defeated if D (cross-cultural transfer fails systematically; the predicted response doesn't follow because some unmodelled stabilizer is operative).

The third sub-category is communicative competence in \textcite{hymes1972}'s sense: the integrated capacity to deploy language as formally possible, feasible, appropriate, and in fact done. Hymes shows that the Chomskyan competence/performance dichotomy is too narrow; what speakers know includes when to speak, when not, what to talk about with whom, when, where, and in what manner. Differential competence within heterogeneous speech communities is the rule, not the exception. Stabilizers are acquisition (children acquire appropriateness rules in the same developmental matrix as grammar), interactive alignment, community transmission of register and address-form conventions, and institutional gatekeeping for some sectors. Observed P (a speaker's performance across multiple contexts) licenses expected Q (predictable performance in unobserved contexts of similar conditioning; predictable judgements about inappropriate uses by others) in unobserved cases through the conditioning-on-context warrant: register, addressee, and discourse community condition appropriateness, and conditioning on these recovers tight prediction from otherwise noisy pooled data. Perturbing the conditioning (eliding register, conflating discourse communities) predicts pooled performance with no recoverable structure. Defeated if D: conditioning on the relevant context variables fails to improve prediction over a pooled baseline (the conditioning-gain test from §\ref{sec:population} returns null for this sub-category), or appropriateness rules don't actually condition speaker performance.

The fourth sub-category is metapragmatic typification, the reflexive level at which speakers and communities recognize forms as indexical of speaker attributes. The apparatus is from linguistic anthropology and the third-wave sociolinguistic tradition. \textcite{agha2005} characterizes \term{enregisterment} as the process whereby distinct forms of speech come to be socially recognized as indexical of speaker attributes through population-level uptake, producing recognizable enregistered voices. \textcite{silverstein2003indexical}'s indexical orders give the layered structure: an n-th order indexical (a form indexing a population, register, or stance) is always available for re-analysis at n+1, with ideological and metapragmatic work driving the re-analysis. \textcite{eckert2008indexical}'s indexical fields capture the constellation of meanings any one form can carry. Stabilizers are reflexive social cognition, looping feedback through speakers' awareness of indexical links, ideological reproduction, and the back-and-forth of convergence and divergence in situated stylistic deployment. Observed P (a form's indexical field, recovered through ethnographic or matched-guise data in community C) licenses expected Q (matched-guise inferences in unobserved community-C members; the form patterning with other registered features when speakers project a recognisable persona; predictable ideological-shift effects on the field as institutional reinforcement varies) in unobserved cases; defeated if D (the indexical field doesn't replicate cross-community, or ideology shifts don't track field shifts).

Modality cross-cuts these sub-categories. Speech, sign, writing, and gesture each support communicative practice with modality-specific affordances and conventions; the emergence of grammar in young sign languages is one of the clearer demonstrations that the apparatus described above is modality-general rather than speech-bound \citep{sandler2006}. The same speech-act, interactional, communicative-competence, and metapragmatic structure operates in signed interaction, with modality-specific instantiations of the conventional infrastructure.

Treating communicative practice as a neighbourhood does what the parallel move at the individual-internal scale (capacity) does. Sub-category individuation makes within-grouping traversals visible as moves needing warrant: a speech-act felicity finding doesn't transfer to a register-uptake claim without showing the metapragmatic step; a conversation-analytic repair pattern doesn't transfer to an appropriateness-rule generalization without engaging the four-sector competence frame; a register-recognition study doesn't transfer to a speech-act-felicity claim without the linking inference that the register conditions felicity. Projection targets get specified at the sub-category grain where the relevant stabilizers actually do their work. The grouping's analytical engine is shared; the homeostasis isn't. The corpus, the treebank, the conversation-analytic transcript, and the LLM output are all analyst's diagnostics for studying communicative practice. Treating any of them \emph{as} communicative practice (as the grouping itself rather than as a representational artefact of researcher choices) substitutes the analyst's diagnostic for the maintenance mechanism. \enquote{The corpus IS the language} reads cross-corpus convergence as evidence about cross-community population convention, when what actually converges is annotation tradition. \enquote{The LLM output IS the language} reads generation traces as data about situated interactional work, when what actually drives them is training-data and tuning artefacts. Section~\ref{sec:slippage} catalogues these alongside the Casten / Pinker case; the neighbourhood's distinctive failure mode is the analyst's-diagnostic-for-maintenance-mechanism substitution that the sub-category discipline above is designed to forestall.

\section{Population-level convention}\label{sec:population}

The category of \mention{language} claims at the population grain, between the individual idiolect and the institutionally-named variety, is what Chomsky calls (and dismisses as artificial) E-language, what \textcite{lewis1969} theorizes as convention, and what Labov's variationist tradition empirically studies. \textcite{santana2016language} rebuts Chomsky's anti-E-language arguments at length, and the rebuttal is largely accepted in this paper. Three sub-categories within the category pursue different research questions and project to different inferences.

The first sub-category is convention in \citeauthor{lewis1969}'s sense: a regularity in the behaviour of members of a population, conditional on a recurrent situation, such that almost everyone conforms, almost everyone expects almost everyone else to conform, almost everyone prefers any one to conform conditional on the others, and at least one alternative regularity would meet the same conditions. Linguistic conventions on this analysis are coordination equilibria: speakers conform to the regularity (use \olang{chien} for dogs in French, \mention{dog} in English, \cjk{犬} in Japanese) because they expect others to and prefer to conform conditional on others conforming. \textcite{lewis1969} opens by calling this a \enquote{platitude that language is ruled by convention} and develops the platitude into an analysis grounded in coordination game theory and Hume's account of common interest. Stabilizers are mutual expectation among community members, communicative success as the payoff function, and iterated interaction producing common knowledge. Observed P (a regularity at the population level) licenses expected Q (the regularity persists under perturbation; alternative coordinations are equally stable in principle but not realized in practice; the regularity tracks common-interest structure of the situation) in unobserved cases; defeated if D (the regularity persists despite individual deviation, suggesting it isn't actually convention-sustained, or the population is too thinly connected for common knowledge to plausibly hold).

The second sub-category is empirical speech-community variation. Labov's foundational work (on Martha's Vineyard's centralized /ay/, on post-vocalic /r/ in New York City department stores, on the orderly heterogeneity that variable rules describe rather than smoothing away) establishes that variation at the population level is systematic, structured, and conditioned by demographic variables (class, gender, age, region) \citep{labov1972, labov2001}. The methodological backbone is sociolinguistic interviewing combined with multivariate analysis; \textcite{tagliamontedarcy2007} and others have extended the apparatus to apparent-time and real-time studies of grammaticalization and change. \textcite{milroy1987} adds network density: tightly connected local networks maintain non-prestige variants and resist change-from-outside, while loose networks accept it. Stabilizers are interactive alignment within the community, identity-marking pressures, network density, and iterated transmission. Observed P (a variable's distribution in interview data) licenses expected Q (similar distribution in unobserved similar speakers; predictable change trajectories under demographic shift; predictable convergence-divergence under network perturbation) in unobserved cases; defeated if D (the distribution doesn't replicate when the community is sampled differently, or conditioning on demographic variables doesn't recover the predicted pattern).

The third sub-category (the third wave of variation studies, in Eckert's periodization) treats variables as stylistic resources carrying ideologically structured social meanings, not as direct reflections of demographic categories. \textcite{eckert2008indexical} argues that variables don't have fixed meanings but inhabit indexical fields: constellations of ideologically related potential meanings, any of which can be activated in situated use. Building on \textcite{silverstein2003indexical}'s indexical orders, the framework treats stylistic moves as ideological work, with speakers positioning themselves with respect to elements of character, persona, and stance that the variable has been enregistered to index. \posscite{eckert2008indexical} reanalysis of \citeauthor{labov1972}'s Martha's Vineyard fishermen makes the move concrete: the centralized /ay/ wasn't simply claiming Vineyarder identity, it was making a claim about what a Vineyarder is, against the encroachment of mainland tourism. Stabilizers are reflexive metapragmatic awareness, the back-and-forth of convergence and divergence, ideological reproduction, and recurrent stylistic deployment. Observed P (a variable's indexical field, recovered through matched-guise or ethnographic data) licenses expected Q (the variable's social meaning in unobserved similar contexts; predictable shifts under ideological pressure) in unobserved cases; defeated if D (the indexical field doesn't replicate cross-community, or ideological-shift predictions fail).

This category carries the methodological commitment that pooled population-level data routinely conflates conditioned systems. Existential agreement gives the category its clearest worked HPC audit. Santana's three-criteria pluralism settles the existence question: English subject-verb agreement exists, is a primary object of study across formal syntax, sociolinguistics, and corpus linguistics, and is reasonably called part of \mention{English}. HPC's discipline picks up where Santana's leaves off, by asking specifically what's maintained at the population level under the maintenance claim.

\emph{Property covariance.} \mention{There's} occurs disproportionately with following plural NPs in informal contexts (\mention{there's two men}); categorical agreement morphology covaries with edited-prose context.

\emph{Stabilizers.} At the population level, two distinct mechanism sets do the work: (a)~interactive alignment in casual conversation lets \mention{there's} stabilize as an invariant presentative particle decoupled from following-NP number, and (b)~institutional editorial gatekeeping in edited prose maintains categorical agreement.

\emph{Perturbation sensitivity.} Weakening either stabilizer predicts patterned decay. Decreasing editorial enforcement (online publishing without copy-editing, for instance) predicts increased \mention{there's two men} in supposedly-edited contexts; loss of register distinction predicts collapse of the two-system pattern toward whichever variant is more common in a given community.

\emph{Projection.} A maintenance claim about \mention{there's} licenses predictions about which registers will tolerate it and which won't, and about which speakers will produce or accept it under which conditioning. The projection target is conditioned, not pooled.

\emph{Defeater.} Cross-corpus failure of the conditioning-gain test would force revision: if rates of \mention{there's two men} versus \mention{there are two men} proved unstructured-noise across registers under properly-controlled comparison, the two-register account would lose its grip. \textcite{krejciHilton2022} document the structured pattern (with a third levelled \mention{there is} variant alongside the two main ones); \textcite{coppock2018double} shows the parallel double-copula case (\mention{the thing is is that\ldots}) is similarly predictable under spontaneous-planning conditioning.

The audit's payoff isn't a contest with Santana's pluralism. Santana's three-criteria framework leaves the maintenance question open; HPC supplies the \term{conditioning-gain test} (whether conditioning on a putative variable improves prediction over the pooled baseline; no gain means no mechanism is sustaining the apparent pattern) that disciplines what counts as a population-maintained pattern. What's maintained here is two register-conditioned systems, not one pooled system. Naming what's maintained is the section's payoff, and the slippage diagnosed in §\ref{sec:slippage} is what happens when pooled data is read as evidence about a single underlying grammar.

\section{Named sociohistorical objects}\label{sec:named}

The category of \mention{language} claims about named varieties (English, French, Quebec French, Oji-Cree, Kʼicheʼ, Japanese, Cantonese, Modern Standard Arabic, Shilluk) sits at the institutional grain rather than the conventional one. What separates it from §\ref{sec:population} is the additional infrastructure: schooling, literacy, codification, legal recognition, language tests, language ideology. Three sub-categories pursue different research questions about the same named objects.

The first sub-category is the named object as institutional artefact. \textcite{haugen1966} catalogued the standardization moves that build named languages: selection of a norm, codification of form, elaboration of function, acceptance by the population. Institutional configurations vary widely across the category. State-backed codification (Modern Standard Arabic, French through the Académie française) sits alongside engineered national-language projects (\mention{Bahasa Indonesia}, modern Israeli Hebrew as a revival project), Indigenous-language community standardization paired with language-rights frameworks (Oji-Cree, Kʼicheʼ), and named varieties whose institutional scaffolding is much lighter (Shilluk is named, recognized, and historically described, but not codified or taught at the scale of the others). National academies, codifying dictionaries, school curricula, language tests, and linguistic-rights legislation are the visible scaffolding where they exist. Stabilizers run from publishing standards through schooling and literacy education to legal and administrative recognition; the intensity varies by variety. Observed P (a feature is institutionally codified as part of named variety V) licenses expected Q (the feature appears reliably in the institutional channels active for V; speakers oriented to V's recognized gatekeepers conform to it) in unobserved cases; defeated if D (the institutional reinforcement is too thin to maintain the predicted patterns, or the feature's distribution doesn't track the institutional channels available).

The second sub-category is the named object as identity-marker. Quebec French is central to Quebecois identity; Kʼicheʼ to Mayan identity in Guatemala; Modern Standard Arabic stands in indexical contrast to vernacular Arabic across the Arab world; Welsh has been the site of revival politics for decades. The metapragmatic machinery here is enregisterment in \citeauthor{agha2005}'s sense: distinct forms of speech come to be socially recognized as indexical of speaker attributes through processes of population-level uptake \citep{agha2005}. Once enregistered, the variety can become emblematic of a recognizable persona, position, or stance. \textcite{irvinegal2000}'s analysis of iconization (a feature is taken to embody a quality of its speakers) and erasure (heterogeneity within the named group is downplayed in the ideological summary) names the ideological machinery that hardens a variable distribution into an identity-emblem. \textcite{hacking1999} captures the feedback: once speakers learn that a form indexes a named variety, their use shifts in response to that knowledge, and the shift reinforces the index. Stabilizers are reflexive metapragmatic awareness, the back-and-forth of convergence and divergence, ideological reproduction, and looping feedback through speakers' awareness of the named-variety identity. Observed P (an indexical association of a named variety with an identity) licenses expected Q (speakers claiming the identity orient to the named variety; community members recognize the indexical association in unobserved cases; ideological shifts reshape the field around the named variety) in unobserved cases; defeated if D (the indexical claim fails to replicate cross-community, or the named variety isn't doing identity-work for the population claimed).

The third sub-category is the named object as descended kind: a sociohistorical lineage tracing through identifiable earlier states. Old English to Middle English to Modern English; Latin to the Romance languages; Common Slavic to its descendants. The methodology is historical-linguistic reconstruction backed by manuscript and textual evidence. Stabilizers are iterated transmission across generations, standardization decisions that anchor snapshots of the named object (literary canons, dictionaries, school texts), and educational continuity that frames the lineage narrative for each cohort. Observed P (a feature in a historical state of the named object) licenses expected Q (predictable diachronic changes in the lineage; predictable correspondences with related languages; predictable reconstructions of unattested earlier states) in unobserved cases; defeated if D (the predicted historical changes don't match the textual record, or the lineage claim turns out to be substantially constructed by post-hoc institutional narrative).

Cross-cutting all three sub-categories is the count-noun deployment of \mention{language} itself. \mention{Cantonese is a language}, \mention{Welsh is a language}, \mention{Jamaican Creole is a language}: the count-noun grammar treats the named object as enumerable, instantiable, and bounded. That count-noun availability is itself a stabilizer at the metapragmatic-ideological grain. It underwrites the bureaucratic infrastructure (census categories, language-test certifications, language-rights legislation) and the educational practices (selecting one norm to teach as \emph{the} standard, distinguishing one named language from a related one in school curricula) that the more visible stabilizers operate through. \citeauthor{haugen1966}'s standardization moves all presuppose count-noun discreteness; \citeauthor{irvinegal2000}'s erasure operates on the assumption that one named language is being demarcated from another; \citeauthor{pennycook2010}'s case against named-language reality is partly an objection to over-reading count-noun ideology as evidence about pre-institutional reality. The count-noun ideology functions in this category the way enregisterment functions in §\ref{sec:communicative}: a reflexive cultural framework that sustains, rather than transparently describes, the named-object pattern. Treating it as one of the category's stabilizers (rather than as pre-institutional reality) keeps the realist commitment located at the right grain and prepares the slippage diagnosis in §\ref{sec:slippage}, which traces what happens when count-noun-licensed recognition data is read as evidence about other elements of the map.

The category's reality has been seriously contested. \textcite{pennycook2010} argues that named languages are inventions, not pre-existing entities the institutions merely label; in the Makoni-and-Pennycook tradition, \enquote{disinventing languages} is the recovery of local practice from the named-object overlay. The critique is partly right. Named-language objects aren't natural kinds awaiting discovery, and the institutional apparatus that names, codifies, teaches, and legally recognizes them is doing constitutive work, not just descriptive labelling. The HPC reframing locates the category's reality at the right grain: a named object is real as an institutionally-stabilized attractor basin, sustained by the schooling, literacy, gatekeeping, and identity-work that maintain it; it isn't real as a pre-existing natural kind, and treating it as one is one of the recurring slippages diagnosed in §\ref{sec:slippage}. The category works at the institutional and identity grain. Reading evidence about it as evidence about cognitive architecture (capacity), individual command (idiolect), or moment-to-moment communicative work (communicative practice) traverses boundaries that the institutional stabilizers don't underwrite.

\section{Abstract formal systems}\label{sec:formal}

The category of \mention{language} claims that treats language as a mathematical or formal object sits at a different grain from the others. Its working desideratum is extensional adequacy: a grammar or formal model captures the language insofar as it generates the right strings or assigns the right meanings, regardless of whether the same operations are performed in any speaker's head or community. Its stabilizers are mathematical-disciplinary (definitional stipulation, mathematical inheritance, and the conventions of the relevant mathematical or computational community), not population-level usage. \textcite{santana2016language} develops this as the abstract ontology, with Platonism (Katz), formalism (a structuralist-style reading), and fictionalism as candidate philosophical positions on what the formal objects are.

The first sub-category is formal-language theory in the sense cited by \textcite{hauser2002}: languages as sets of strings, grammars as finite devices that generate or recognize those sets, and the Chomsky hierarchy that ranks grammars by descriptive power. Natural languages are at least mildly context-sensitive: \citeauthor{pinkerjackendoff2005} cite the Swiss-German cross-serial dependencies and Bambara reduplication results that put natural languages above context-free in the standard hierarchy. \textcite{jagerrogers2012} review the refinements relevant to cognitive science: mildly context-sensitive grammars (Tree-Adjoining Grammar, Combinatory Categorial Grammar, minimalist grammars) above context-free, and the sub-regular hierarchy below regular for local-pattern phenomena. Stabilizers are mathematical inheritance and the disciplinary conventions of formal-language theory. Observed P (a natural-language phenomenon like centre embedding or cross-serial dependencies) licenses expected Q (the language is at least at hierarchy level X; the cognitive mechanism recognizing it has to have at least the corresponding power) in unobserved cases; defeated if D (the formal-language-theoretic characterization turns out to be inadequate to phenomena that exceed it, or doesn't actually correspond to the cognitive mechanism in question).

The second sub-category is formal semantics in the Montague tradition. Meanings are model-theoretic objects (functions, sets, possible-world structures) composed by rules that map syntactic structure to semantic value. The compositionality move treats the meaning of a complex expression as determined by the meanings of its parts and the way they're combined; type-theoretic and intensional developments extend the apparatus to quantification, modality, and dynamic phenomena. Stabilizers are model-theoretic and type-theoretic inheritance. Observed P (a semantic property of expressions in some test cases) licenses expected Q (the same property in unobserved cases of similar structure; predictable inference patterns; cross-linguistic correspondences) in unobserved cases through compositional-semantic warrant: the meaning of a complex expression is determined by the meanings of its parts and how they're combined, so structural similarity carries the inference. Perturbing the compositional architecture (introducing non-compositional idioms, pragmatic enrichment, or contextually shifty meanings) predicts patterned decoupling of structural similarity from meaning prediction. Defeated if D: the formal-semantic characterization fails on actual meaning judgements; cross-linguistic universals don't replicate; phenomena resist clean compositional analysis at scale.

The third sub-category is mathematical-Platonist linguistics in \textcite{katz1981}'s sense: linguistic objects are abstract entities analogous to numbers in mathematical Platonism. The grammar describes those abstract objects; extensional adequacy is achievable without commitment to psychological reality. Stabilizers are philosophical-Platonist tradition and disciplinary practice in the (smaller) Platonist-leaning subset of formal linguistics. Observed P (a structural pattern in linguistic data) licenses expected Q (the same pattern in unattested cases of similar formal structure with the necessity characteristic of mathematical truth; counterfactual stability under variation in psychological and sociological inputs) in unobserved cases; defeated if D (alternative ontologies~-- psychological, social, conventionalist, fictionalist~-- explain the patterns equally well, leaving Platonism to do no extra explanatory work).

The fourth sub-category is computational and NLP working usage: Carnap-style methodological idealization at the engineering scale. Parsers, generators, machine-translation systems, and (now) large language models all treat formal models of language as the working object, with success measured by performance benchmarks rather than psychological or sociological fidelity. Stabilizers are engineering tradition, computational-linguistic disciplinary practice, and the cumulative apparatus of mathematical inheritance from sub-categories 1 and 2. Observed P (a system's performance on a class of tasks) licenses expected Q (similar performance on similar tasks; predictable transfer; predictable failures outside the class) in unobserved cases; defeated if D (transfer fails in ways the model doesn't predict, or the model's success comes from properties unrelated to the formal characterization).

The category's typical traversal is to capacity. \citeauthor{hauser2002}'s Hypothesis~3, that the faculty of language in the narrow sense (FLN) consists of recursion plus the interfaces with sensory-motor and conceptual-intentional systems, is a worked example. Recursion is well-defined as a formal-language-theoretic property (sub-category~1). FLN, in \citeauthor{hauser2002}'s sense, is a cognitive-architectural posit (capacity sub-category 1 in §\ref{sec:capacity}). \citeauthor{pinkerjackendoff2005} argue that the recursion-only thesis transfers a §\ref{sec:formal} sub-category~1 characterization to a §\ref{sec:capacity} sub-category~1 claim without doing the work that bridge would need: phonology has discrete infinity but isn't technically recursive; words encode language-specific syntactic information not captured by recursion alone; case, agreement, and most of the syntactic apparatus beyond hierarchical phrase grouping are language-specific. The disagreement is partly empirical, partly methodological. The HPC discipline names the methodological half: a formal-language-theoretic property is the kind of thing that's stabilized by mathematical inheritance, and a faculty-architecture claim is the kind of thing that's stabilized by genetic and developmental machinery; the two stabilizer kinds don't underwrite each other's inferences, and §\ref{sec:slippage} catalogues this category-to-neighbourhood move alongside the Casten / Pinker case.

\section{Slippage}\label{sec:slippage}

The taxonomy from §§\ref{sec:capacity}--\ref{sec:formal} earns its keep when it does diagnostic work that Santana-style pluralism leaves underspecified. \enquote{Slippage} is the recurring move where evidence licensed by the stabilizers of one element gets read as evidence about another that the term \mention{language} also covers, whether across the map (category to category, category to neighbourhood, or vice versa) or within a neighbourhood (between sub-categories that don't share a stabilizer set). HPC supplies the discipline: each category's stabilizers underwrite a specific projection target; each neighbourhood's sub-categories do likewise at their own grain; and when stabilizers don't transfer across the relevant boundary, the inference doesn't either. This section walks through the Casten / Pinker case in detail, catalogues other recurrent slippage classes, and closes with the conditions under which evidence \emph{can} legitimately travel across the map.

The Casten / Pinker case from §\ref{sec:intro} is the worked instance. \textcite{casten2026} measure F1, an aggregate developmental phenotype anchored by sentence repetition in 350 children and replicated with phenotypically coarser measures in over 100,000 participants. They report a polygenic score restricted to human-ancestor quickly-evolved regions (HAQERs) with $\beta = 0.18$ on F1, no association with nonverbal IQ, and HAQER-like sequence similarity correlated with vocal-learning ability across 49 vocal-learning and 121 non-vocal-learner mammals. Pinker, citing the paper as vindication of \textcite{pinkerbloom1990}, reads it as evidence for a uniquely human modular language faculty.

Figure~\ref{fig:slippage} schematizes projection warrant in the case, with the faculty target deliberately split. F1 is the measured aggregate, anchored by contributors that include working memory, attention, lexical access, motor speech, and grammatical processing. HAQER CP-PGS is reported as the upstream genomic anchor for F1 within the developmental sub-category (sub-category 3 in §\ref{sec:capacity}'s neighbourhood). HAQER-like sequence similarity correlates with vocal-learning phenotype across nonprimate mammals within the comparative-biology sub-category (sub-category 4). Solid links in the schematic mark evidence that is locally projectible within the stabilizer profile of each sub-category. The dashed link from F1 to the Pinker-Bloom broad faculty marks the inference Pinker's tweet actually makes: a complex genetic adaptation distinct from general cognition \citep{pinkerbloom1990}. The further dashed link from broad faculty to HCF narrow FLN \citep{hauser2002} marks an additional, more specific, narrowing step (recursion plus the interfaces with sensory-motor and conceptual-intentional systems). Pinker's tweet doesn't take the second step explicitly, but the same evidence base often gets used to license both claims at once. Splitting them is part of the discipline: F1's stabilizers (developmental, comparative-biology) don't underwrite the broad-faculty claim, and the broad-faculty claim doesn't underwrite the narrow-FLN claim either. Those are separate inference burdens that the Pinker--Bloom thesis and the HCF programme each take on, and which \citeauthor{pinkerjackendoff2005} contest in different ways.

The issue isn't whether F1 causes the faculty; it's whether evidence about developmental verbal performance and comparative vocal-learning regulatory architecture licenses claims about a uniquely human computational faculty, whatever its specific characterization. It doesn't, on the standard bridge-principle conditions: HAQER's stabilizers are regulatory and shared with mammalian vocal learners, the faculty posit's stabilizers are contested between UG-style genetic endowment and convergent product of multiple selections, and neither story routes through F1. If HAQER-like architecture is shared with cetaceans, pinnipeds, and bats, what F1 anchors is a capacity humans share with nonprimate mammalian vocal learners, not a capacity uniquely human. The slippage is projection along an unsupported warrant chain, not a causal claim about F1 producing the faculty.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  >={Stealth[length=2mm]},
  font=\small,
  box/.style={draw, rounded corners=2pt, inner sep=4pt, align=center, font=\footnotesize, text width=32mm, minimum height=10mm},
  measure/.style={box, fill=black!7},
  cluster/.style={draw=gray, rounded corners=4pt, inner xsep=7pt, inner ysep=8pt, dashed},
  clusterlabel/.style={font=\scriptsize\itshape, align=left},
]
% Sub-category 4 nodes
\node[clusterlabel, anchor=west] (label4) at (-51mm,0mm) {Sub-category 4 -- comparative biology};
\node[box] (haqer-cross) at (-24mm,-11mm) {HAQER-like sequences\\(across mammals)};
\node[box] (vocal) at (24mm,-11mm) {Vocal-learning phenotype\\(49 species)};

% Sub-category 3 nodes
\node[clusterlabel, anchor=west] (label3) at (-64mm,-29mm) {Sub-category 3 -- developmental};
\node[box] (haqer-pgs) at (-48mm,-41mm) {HAQER CP-PGS};
\node[measure] (f1) at (0mm,-41mm) {F1 phenotype\\(sentence repetition)};
\node[box] (contrib) at (48mm,-41mm) {Working memory\\Attention\\Lexical access\\Motor speech\\Grammar};

% Sub-category 1 nodes (split into broad / narrow)
\node[clusterlabel, anchor=west] (label1) at (-64mm,-66mm) {Sub-category 1 -- faculty architecture};
\node[box] (fln-broad) at (-22mm,-82mm) {Pinker-Bloom\\broad faculty (1990):\\complex genetic adaptation\\distinct from general cognition};
\node[box] (fln-narrow) at (22mm,-82mm) {HCF narrow FLN (2002):\\recursion + interfaces};

% Solid edges (supported)
\draw[->] (haqer-cross) -- (vocal);
\draw[->] (haqer-pgs) -- (f1);
\draw[->] (contrib) -- (f1);

% Dashed edges (unsupported)
\draw[->, dashed, thick] (f1) -- node[left, align=center, font=\scriptsize\itshape, pos=0.55] {Pinker tweet:\\unsupported\\bridge} (fln-broad);
\draw[->, dashed, thick] (fln-broad) -- node[above, align=center, font=\scriptsize\itshape] {HCF 2002\\further narrowing} (fln-narrow);

% Sub-category boundaries
\begin{scope}[on background layer]
\node[cluster, fit={(label4)(haqer-cross)(vocal)}] {};
\node[cluster, fit={(label3)(haqer-pgs)(f1)(contrib)}] {};
\node[cluster, fit={(label1)(fln-broad)(fln-narrow)}] {};
\end{scope}
\end{tikzpicture}
\caption[Projection-warrant schematic for the Casten / Pinker slippage]{Projection-warrant schematic for the Casten / Pinker slippage, with the faculty target split into Pinker-Bloom broad and HCF narrow versions. Solid links mark locally warranted evidential or projective links within sub-categories. The dashed link from F1 to the broad faculty marks the inference Pinker's tweet makes; the dashed link from broad to narrow marks the further narrowing step \citep{hauser2002} that the same evidence is sometimes claimed to license. None of the three bridge-principle conditions (linked stabilizers, projection-target subsumption, auxiliary linkage evidence) is satisfied for either dashed link. The schematic is local to this biocognitive case; sociocultural slippages involving uptake, recognition, and institutional ratification need feedback structure this kind of diagram doesn't supply.}
\label{fig:slippage}
\end{figure}

A second worked case shifts the slippage to sociocultural ground at full audit length. The named sociohistorical object \mention{Cantonese} is recognized by census instruments (Hong Kong, Statistics Canada, U.S.~Census Bureau), schooled in Hong Kong and parts of Guangdong, codified in dictionaries and language-test syllabi, recognized in some language-rights frameworks, and enregistered as indexical of Cantonese-speaker identity across the Sinosphere. The named object earns category status at the institutional grain (§\ref{sec:named}'s audit). Census recognition data is the route through which the slippage to population convention appears, and the audit walks the same five elements as §\ref{sec:population}'s existential-agreement case.

\emph{Property covariance.} Within the named-object frame, features cluster: census-claimed Cantonese-speaker identity, school exposure to Cantonese, dictionary access, language-rights-claim engagement, and identity-marking work co-occur in identifiable populations across the Sinosphere and in the diaspora. The covariance is real at the institutional grain.

\emph{Stabilizers.} Per §\ref{sec:named}'s three sub-categories plus the count-noun cross-cut: institutional ratification (Basic Law treatment in Hong Kong; Putonghua-policy positioning in mainland China; varying institutional channels in diaspora communities), identity-marking through enregisterment \citep{agha2005}, looping feedback through speakers' awareness of their Cantonese-speaker identity \citep{hacking1999}, and the count-noun ideology that licenses bounding and enumeration. The stabilizers are institutional and metapragmatic, not interactional-coordinative.

\emph{Perturbation sensitivity.} Shift the institutional regime (the 1997 Hong Kong handover; mainland Putonghua-promotion campaigns; changes in Canadian census language categories) and the named-object category reconstitutes at a different intensity but holds. A putative \mention{Cantonese-as-population-convention} category (a single Lewis-style coordinative regularity uniting the population) doesn't reconstitute under the same perturbations, because it wasn't maintained as one coordinated regularity in the first place; what reconstitutes is named-object recognition at a regime-specific intensity.

\emph{Projection.} The named-object category projects to institutional reach (where Cantonese is taught, certified, recognized) and to identity-recognition (who claims Cantonese-speaker status, what indexical work the claim does). It does not project to population-convention rates, individual-cognitive-system uniformity, or speech-act felicity conditions in interaction.

\emph{Defeater.} The named-object inference would be defeated by a count that holds without institutional ratification, or one that fails to track speakers' own identity-claims. Neither obtains: institutional ratification and identity-claim are what the count is tracking.

The slippage is reading recognition data as evidence about population convention. \enquote{X million speakers of Cantonese} gets read as evidence about Cantonese-as-population-convention, as if a single Lewis-style coordinative regularity unites the population. What census instruments actually measure is institutional recognition filtered through enregistered identity-claims and the count-noun ideology of §\ref{sec:named}. The named-sociohistorical category's stabilizers (institutional ratification, identity-marking, looping feedback, count-noun ideology) generate the recognition data; the population-convention category's stabilizers (interactive alignment, transmission, network density) would have to generate any convention claim; the two stabilizer sets don't underwrite each other's inferences. \citeauthor{pennycook2010}'s case against named-language reality lands here: treating count-noun ideology as transparent description of pre-institutional regularity is exactly the move HPC blocks at the named-object-to-population-convention boundary.

None of the three bridge-principle conditions is satisfied for the recognition-to-convention inference. Linked stabilizers: institutional channels (schools, censuses, gatekeeping) don't causally contribute to interactive alignment in any general way; they shape what speakers report doing, not the population-level coordinative regularity. Projection-target subsumption: the named-object's institutional-reach target doesn't include the population-convention category's coordinative-regularity target. Auxiliary linkage evidence: a successful bridge would require independent evidence that institutional ratification predicts interactional coordination at the population grain. That evidence is sometimes available (standard-variety schooling can shift convention rates over generations) but has to be supplied case-by-case, not assumed. Without the bridge, the slippage is the sociocultural counterpart of the biocognitive Casten / Pinker case: warrant-chain expansion across element boundaries, with element-specific stabilizers doing constitutive work that the reading erases.

The same pattern recurs across other element pairs. \emph{Idiolect into population convention}: Chomskyan I-language data is sometimes presented as evidence about (or against) E-language regularities, and population-level variation studies sometimes get read as direct evidence about individual mental grammars; the §\ref{sec:idiolect} neighbourhood's sub-category stabilizers and §\ref{sec:population}'s category stabilizers operate at different grains (HPC-book Levels~1 versus 2--3). \emph{Named object into convention}: a feature institutionally codified as part of a named variety is taken to be a population-level convention, ignoring that the institutional channels (schooling, codification, gatekeeping) have different reach than the conventional regularities they overlap with. \emph{Formal system into empirical object}: formal-language-theoretic results about the abstract system get read as claims about the cognitive faculty that produces utterances, with the gap between mathematical-disciplinary stabilizers and individual-cognitive stabilizers smoothed over. The HCF Hypothesis 3 \enquote{recursion only} thesis is partly this slippage: a formal-language-theoretic property characterized in §\ref{sec:formal}'s category gets transferred to an architectural claim about §\ref{sec:capacity}'s sub-category 1.

A second class is operationalization slippage: the analyst's diagnostic for an element substituted for the element itself. \enquote{The corpus IS the language} reads cross-corpus convergence (a sign of shared annotation conventions) as evidence about cross-community population convention. \enquote{Competence IS the language} reads idealized acceptability judgements as direct windows onto idiolect or i-language. \enquote{The treebank IS the grammar} reads parsed structures as claims about the underlying system rather than about the annotation scheme. \enquote{The LLM output IS the language} reads generation traces as data about communicative practice or capacity, conflating training-data-and-tuning artifacts with situated interactional work. In every case the analyst's diagnostic is one of the three uses of \enquote{mechanism} in HPC analysis, alongside maintenance mechanisms (population-level) and speaker mechanisms (individual production), and treating the diagnostic as the maintenance mechanism is exactly the slippage that the three-way distinction is meant to forestall.

Grammaticality, taken as a worked locus, exhibits a related family of traversals across the map. Acceptability ratings are sometimes treated as direct evidence about grammaticality, conflating a feeling-dimension diagnostic (subjective response to processing, social, and prescriptive cues) with the licensing dimension (a form-meaning pairing's status in a conditioned population's convention). Processing difficulty (centre-embedding, garden-path effects) is sometimes read as evidence about grammatical status, transferring a capacity-neighbourhood phenomenon to a population-convention claim without bridge. Corpus frequency is sometimes treated as automatic evidence of grammatical status, although high-frequency errors fail to entrench and rare-but-licensed forms persist. Formal-grammar generation is sometimes treated as evidence about community licensing, transferring a formal-language-theoretic characterization to a population-convention claim. Named-language norms (\enquote{standard English says X}) are sometimes presented as evidence about population-level convention, transferring an institutional claim to a convention claim. Each is an element-boundary crossing of the kind diagnosed above; together they form a methodological family that the bridge principles below adjudicate. Treating grammaticality itself as an HPC regime instantiated by conditioned linguistic ecologies, rather than as a single global kind, lines up with the map: grammaticality lives primarily in §\ref{sec:population}'s population-convention category, with bridges to capacity (processing sub-category), communicative practice (acceptability as situated response), named sociohistorical objects (institutional norms), and abstract formal systems (formal grammars).

Slippage diagnosis would be hollow without a positive account of when evidence \emph{can} travel across the map. Three conditions earn the warrant. The first is demonstrably linked stabilizers: when the maintenance mechanisms of element A causally contribute to those of element B, evidence about A bears on B with the strength of the demonstration. Population-level convention shapes individual idiolect through transmission and entrenchment; once that linkage is established, population data informs idiolect inferences with appropriate scope. The second is projection-target subsumption: when element A's projection target includes element B's, evidence supporting A's projection licenses (weaker) inferences about B. Capacity sub-category 4 (comparative-biology) underwrites some of sub-category 3 (developmental) in the special cases where regulatory architecture predicts developmental outcome. The third is auxiliary linkage evidence: when independent evidence ties element A's stabilizers to element B's outputs in particular cases, the auxiliary evidence carries the inference and the boundary is honored rather than crossed. Bridge principles do for cross-element traversal what \citeauthor{santana2016language}'s no-isolationism stricture demands: the elements interact, but the interaction has to be argued, not assumed.

The diagnostic and the bridge principles work as a pair. Naming the slippage in HPC terms makes the move legible as a methodological failure rather than a rhetorical glide. Naming the conditions for legitimate traversal protects pluralism from collapsing into isolated parishes. Together they let \mention{language} do the cross-disciplinary work the umbrella label is for, without smuggling unearned inferences across map elements that don't share the relevant stabilizers.

\section{The umbrella label}\label{sec:umbrella}

The umbrella label \mention{language} keeps its place across the language sciences because it does coordinative work no element-label alone can do. Linguists, philosophers of mind, anthropologists, sociolinguists, computer scientists, and educators converge on the term because it gestures at the field of inquiry they all participate in, even when their specific categories and projection targets diverge. The convergence is genuinely useful: cross-disciplinary work depends on shared vocabulary, and replacing \mention{language} with six element-labels in every cross-field context would be both impractical and rhetorically self-defeating.

The coordinative role doesn't license a realist commitment, though. For the umbrella to be a higher-order HPC category, it would need its own projection target (a class of inferences licensed by the umbrella that aren't already licensed by some individual element) and its own higher-level stabilizers. Neither is in evidence. The map's projection targets concern faculty architecture (or, more precisely, capacity sub-category architectures), individual command (across idiolect's sub-categories), situated communicative practice (across its sub-categories), population-level convention, named-variety institutional infrastructure, and formal-system extensional adequacy. None of these has the umbrella as its target; the umbrella names the (shifting, contested, partial) overlap among them. The stabilizers are the genetic-developmental, individual-cognitive, dyadic-interactional, conventional-population, institutional, and mathematical-disciplinary mechanisms catalogued in §§\ref{sec:capacity}--\ref{sec:formal}. No higher-level stabilizer maintains the umbrella as a unified entity; what unites the elements is the shared label and the shared inquiry, not a shared mechanism set.

The umbrella is a coordinative term, doing the work of cross-field inquiry. Treating it as a higher-order kind would invite exactly the inflation a pluralist position has to resist. The categories and neighbourhoods are the load-bearing units. The umbrella does the work of letting them sit on the same conference programme.

\section{Conclusion}\label{sec:conclusion}

The question \mention{What is language?} has been a productive provocation across the language sciences for at least a century. As a question to be answered straightforwardly, it has also been an obstacle: positive answers in one category get presented as the answer that supersedes others, and reception traffic across category boundaries goes unnoticed. Three replacement questions do better.

The first is which element of the map is in play. \mention{Language} as deployed in a given research context picks out one or more of the six elements mapped above (or, for groupings that the test of §\ref{sec:pluralism} resolves into a neighbourhood, one or more sub-categories within it), or, for new phenomena, an element yet to be added to the taxonomy. Naming what's in play is a methodological commitment: it specifies the stabilizers under discussion and the kinds of inferences the element's mechanisms underwrite.

The second is what projection target the classification licenses. A claim about capacity sub-category 4 (comparative biology) licenses inferences about cross-species regulatory architecture. A claim about population-level convention licenses inferences about community coordination patterns under conditioning. A claim about a named sociohistorical object licenses inferences about institutional reach. Conflating projection targets is the engine of the slippages catalogued in §\ref{sec:slippage}.

The third is what slippage risks attend the use of the word in a given case. Some elements are adjacent enough that traversal is sometimes warranted, on the bridge principles of §\ref{sec:slippage}: demonstrably linked stabilizers, projection-target subsumption, or auxiliary linkage evidence. Others have stabilizer profiles that make traversal across them very rarely warranted. Naming the risks, and the conditions for warrant, protects pluralism from collapsing either into isolated parishes or into unearned cross-element inference.

The taxonomy is provisional. New elements are admissible if they pass the projectibility and homeostasis tests; existing elements can be split, merged, or demoted to neighbourhoods, as capacity, idiolect, and communicative practice already have been. What working scientists do when they say \mention{language} should be diagnosable. The map is the diagnostic instrument, and the projectibility-and-homeostasis discipline is its calibration.

\section*{Acknowledgements}

This paper was drafted and revised with assistance from Anthropic Claude and OpenAI Codex (GPT-5). All content has been reviewed and revised by the author, who takes full responsibility for the final text.

\printbibliography

\end{document}
